Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-07-27 Manuscript/proto-manuscript.tex
MS: updated section Comparison with experiments

Here we apply the electronic continuum correction to lipid bilayers to accurately 
describe the lipid headgroup response to Na+ and Ca2+ concentrations \cite{catte16}. 
We use the Lipid14~\cite{dickson14} force field parameters as a starting point,
because they give the most realistic headgroup response with added cations 
and relatively realistic glycerol backbone and headgroup structures
when compared with other state of the art lipid models \cite{botan15,catte16}.
The partial atomic charges in Lipid14 were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum.
If the charges are obtained by using RESP in an implicit solvent, 
they vary the most for the polar moieties \cite{maciejewski14}. 
In order to implicitly capture solvent induced polarization 
in models with fixed charges, 
we shall use the average of partial charges 
from both environments (i.e. vaccum and solvent),
so called implicitly polarized charges (IpolQ) \cite{Cerutti2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IpolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway, I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

@@

Mere scaling of partial charges in the modified region by the factor $f_q$ reduces the 
area per molecule to ??\todo{find the value}, which is significantly smaller than the
experimental value (\cite{}) and the original Lipid14 values (\cite{}).  
The decrease of area was speculated 
\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
to arise from reduced hydration
of headgroup due to the lower polarity of molecules with scaled charges.
The hydration can be increased by decreasing the radius of atoms
by reducing the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
This allows water molecules to approach closer to lipid atoms and
have stonger electrostatic interactions with them.
\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Indeed, by reducing $\sigma$ with a scaling factor $f_\sigma = 0.89$ 
we increased the area per molecule
to a level close to experimental and original Lipid14 values. 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison to experimental data}
Structures sampled by individual lipid molecules were compared
to experimental data by using C-H bond order parameters \cite{ollila16}
\begin{equation}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between C-H bond and membrane
normal and average is taken over all sampled configurations.
Bilayer dimensions were compared to experiments by using the
scattering form factor \cite{ollila16}
\todo{Add equation}.

Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}. More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative \cite{seelig87}.

%\todo{ongoing,Actual concentration of cations in simulation has yet to be estimated. 
%If it varies too much from the nominal concentration, I may need to tweak the scaling 
%factors, $f_q$ or only $f_\sigma$, to accomodate it. However, it is very unlikely, response 
%to the surfactant \ce{DHMDMAB} is OK. Big patches with loads of solvent are running at the 
%moment to guide me on the possible finite-size errors and this conc-error.
The used experimental data reports order parameters as a function of
equilibrium cation concentration in bulk solvent \cite{akutsu81,altenbach84}.
The box size in simulations was adjusted such that ion conceration reaches 
a clear bulk plateau, which then gives the concentration that corresponds to the
one reported in experiments. The concentrations in units of mol/l were then 
determined as
\begin{equation}
  C_{eq}=\frac{C_{plateau}}{0.602},
\end{equation}
where plateau concentration is the number density in unist of 1/(nm)$^3$.
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }

>>>>

Here we apply the electronic continuum correction to lipid bilayers to accurately 
describe the lipid headgroup response to Na+ and Ca2+ concentrations \cite{catte16}. 
We use the Lipid14~\cite{dickson14} force field parameters as a starting point,
because they give the most realistic headgroup response with added cations 
and relatively realistic glycerol backbone and headgroup structures
when compared with other state of the art lipid models \cite{botan15,catte16}.
The partial atomic charges in Lipid14 were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum.
If the charges are obtained by using RESP in an implicit solvent, 
they vary the most for the polar moieties \cite{maciejewski14}. 
In order to implicitly capture solvent induced polarization 
in models with fixed charges, 
we shall use the average of partial charges 
from both environments (i.e. vaccum and solvent),
so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway, I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

@@

Mere scaling of partial charges in the modified region by the factor $f_q$ reduces the 
area per molecule to ??\todo{find the value}, which is significantly smaller than the
experimental value (\cite{}) and the original Lipid14 values (\cite{})\todo{Add the values}.  
The decrease of area was speculated 
\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
to arise from reduced hydration
of headgroup due to the lower polarity of molecules with scaled charges.
The hydration can be increased by decreasing the radius of atoms
by reducing the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
This allows water molecules to approach closer to lipid atoms and
have stonger electrostatic interactions with them.
\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Indeed, by reducing $\sigma$ with a scaling factor $f_\sigma = 0.89$ 
we increased the area per molecule
to a level close to experimental and original Lipid14 values. 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison to experimental data}
Structures sampled by individual lipid molecules in simulations were compared
to experimental data by using C-H bond order parameters \cite{ollila16}
\begin{equation}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between C-H bond and membrane
normal and average is taken over all sampled configurations.

Bilayer dimensions were compared to experiments by using the
scattering form factor \cite{ollila16}
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}. More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}.

%\todo{ongoing,Actual concentration of cations in simulation has yet to be estimated. 
%If it varies too much from the nominal concentration, I may need to tweak the scaling 
%factors, $f_q$ or only $f_\sigma$, to accomodate it. However, it is very unlikely, response 
%to the surfactant \ce{DHMDMAB} is OK. Big patches with loads of solvent are running at the 
%moment to guide me on the possible finite-size errors and this conc-error.
The used experimental data report order parameters as a function of
equilibrium cation concentration in the bulk solvent \cite{akutsu81,altenbach84}.
Such a condition is reached in simulations by adjusting
the simulation box size to dimensions large enough
that ion conceration reaches a clear plateau in the bulk solvent.
The concentrations in the units of $\mathrm{mol/l}$ were then 
determined as
\begin{equation}
  C_{eq}=\frac{C_{plateau}}{0.602},
\end{equation}
where plateau concentration is the number density in the units of $\mathrm{nm}^{-3}$.
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }
